\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-08-01 papers/methods/paper.tex
More changes from Tam before refactor

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, increased treatment of systematics, and analytic correction of selection effects, including systematic uncertainty on those corrections. We also desire this to be more computationally efficient than prior works, such that cosmological results from thousands of supernovae are obtainable in the order of hours, rather than days. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017}. The primary challenge of fitting hierarchical models is their large number of fit parameters, and Stan, which uses automatic differentiation and the no-U-turn Sampler (NUTS) (a variant of Hamiltonian Monte Carlo), allows us to efficiently sample high dimensional parameter space.

@@

As we are focused on the spectroscopically confirmed supernovae for this iteration of the method, we assume the observed redshift $\hat{z}$ is the true redshift $z$ such that $P(\hat{z}|z) = \delta(\hat{z} - z)$. Potential sources of redshift error (such as peculiar velocities) are taken into account not via uncertainty on redshift (which is technically challenging to implement) but instead uncertainty on distance modulus. This is discussed further in Section \ref{sec:systreat}. Similarly, we take the mass probability estimate $\hat{m}$ as correct, and do not model a latent variable to represent uncertainty on the probability estimate. One of the strengths of this model (and the {\rubin} analysis) is that for future data sets where supernovae have been classified photometrically, and we expect some misclassification and misidentification of the host galaxies, those can naturally be modelled and taken into account.

@@

For the moment, let us consider a single supernova, and represent its true parameters as $\theta$. The underlying supernova model and intrinsic dispersion model gives the probability of $p(y|\theta)$ for obtaining light curve data $y$. As mentioned earlier, we work with summary statistics -- numbers output from a light curve fitter given the input $y$. The function $f$ describes the action of the fitter $\lbrace \hat{m}_B, \hat{x}_1., \hat{c} \rbrace = f(y)$. For notational simplicity, let us denote $\eta \equiv \lbrace m_B, x_1, c \rbrace$. The model prediction for our derived data is

@@

{\green Unfortunately, the assumption of Gaussian uncertainty is incorrect - the errors reported by the light curve fitting process on $m_B$, $x_1$ and $c$ are not accurate. The intrinsic supernovae population is commonly described in terms of coherent magnitude dispersion and in chromatic dispersion, where respectively some supernovae are uniformly brighter or dimmer than the population mean, and the spectral energy density of the supernovae varies non-uniformly as a function of wavelength. The presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics reported from light curve fitting are under-reported, not Gaussian and equation \eqref{eq:pop} is violated. We address this in two ways - increasing the reported uncertainty, and allowing for a global shift in observed colour.}

We model the extra dispersion only in colour, and do so by adding extra independent uncertainty on the colour observation. As shown in \citep{Kessler2013}, the extra colour dispersion shows heavy redshift dependence. As the extra dispersion may arise from underlying supernova physics, rather than Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. The $\kappa$ parameters are tightly constrained as they are highly degenerate with the width of the intrinsic colour population $\sigma_c$, and the specific value of this prior was determined from running many iterations of simulation fitting with differing prior strengths on our simple data set with simulated extra dispersion according to the results of \citet{Kessler2013}. The value $0.002$ can be varied by a factor of approximately three in either direction whilst giving stable results, with larger values (more than a factor of three) for the prior resulted in incorrect determination of $\sigma_c$, whilst smaller values resulted in being unable to capture the extra dispersion  Each survey has its own set of $\kappa$ parameters. As such, our combined covariance on the observation $\hat{\eta}$ is given by $\mathcal{C}_{\rm tot} = \mathcal{C} + {\rm DiagMatrix}\left[0, 0, (\kappa_0 + \kappa_1 z)^2\right]$.

@@

Our second way of addressing the issues with uncertainty are to use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift. {\green This redshift-dependent array is added to our data-set and is included in our modelling by allowing a controlled subtraction of this shift from our observed data, where the amount of shift subtracted is dictated by parameter $s_c$. Specifically, allow the observed colour to shift by $c_{\rm new} = c_{\rm biased} - s_c \times {\rm bias}(z)$. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a global shift in supernovae colour across all surveys. We place a uniform prior on $s_c$.} With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations. The scatter models, sourced from \citet{Guy2010} and \citet{Chotard2011} are discussed in detail in Section \ref{sec:simdes}. The maximum shift was found to using the intrinsic scatter model from \citet{Chotard2011}. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

	% Plates
	\plated[thick] {sn} {(obssum)(zobs)(mobs)(pobs)(sum)(z)(m)(mu)} {$\forall\ \rm{SN}$} ;
	\plated[thick] {survey} {(sn)(pop1)} {\\$\forall\ \rm{Survey}$} ;
	\end{tikzpicture}
	\caption{Probabilistic graphical model for our likelihood. Double-lined nodes represent observed variables, diamond nodes represent deterministic variables, and standard nodes represent fit variables. The SN box represents observed variables and latent variables for each individual supernova, whilst the survey box represents survey-specific variables, which in general describe the supernova population for the survey and the systematics associated with it.}
	\label{fig:pgm}
 \end{figure}

>>>>

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, increased treatment of systematics, and analytic correction of selection effects, including systematic uncertainty on those corrections. We also desire this to be more computationally efficient than prior works, such that cosmological results from thousands of supernovae are obtainable in the order of hours, rather than days. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in Stan \citep{Carpenter2017, StanDevelopmentTeam2017}. The primary challenge of fitting hierarchical models is their large number of fit parameters, and Stan, which uses automatic differentiation and the no-U-turn Sampler (NUTS, a variant of Hamiltonian Monte Carlo), allows us to efficiently sample high dimensional parameter space.

@@

As we are focused on the spectroscopically confirmed supernovae for this iteration of the method, we assume the observed redshift $\hat{z}$ is the true redshift $z$ such that $P(\hat{z}|z) = \delta(\hat{z} - z)$. Potential sources of redshift error (such as peculiar velocities) are taken into account not via uncertainty on redshift (which is technically challenging to implement) but instead uncertainty on distance modulus. This is discussed further in Section \ref{sec:systreat}. Similarly, we take the mass probability estimate $\hat{m}$ as correct, and do not model a latent variable to represent uncertainty on the probability estimate. One of the strengths of this model (and the {\rubin} analysis) is that for future data sets where supernovae have been classified photometrically, and we expect some misclassification and misidentification of the host galaxies, those can naturally be modelled and taken into account by introducing secondary populations.

@@

For the moment, let us consider a single supernova, and represent its true parameters as $\theta$. The underlying supernova model and intrinsic dispersion model gives the probability of $p(y|\theta)$ for obtaining light curve data $y$. As mentioned earlier, we work with summary statistics -- numbers output from a light curve fitter given the input $y$. The function $f$ describes the action of the fitter $\lbrace \hat{m}_B, \hat{x}_1, \hat{c} \rbrace = f(y)$. For notational simplicity, let us denote $\eta \equiv \lbrace m_B, x_1, c \rbrace$. The model prediction for our derived data is

@@

{\green Unfortunately, the assumption of Gaussian uncertainty is incorrect - the errors reported by the light curve fitting process on $m_B$, $x_1$ and $c$ are not accurate. The intrinsic supernovae population is commonly described in terms of coherent magnitude dispersion and in chromatic dispersion, where respectively some supernovae are uniformly brighter or dimmer than the population mean, and the spectral energy density of the supernovae varies non-uniformly as a function of wavelength. The presence of chromatic smearing in the intrinsic supernova population \citep{Guy2010, Chotard2011} means the summary statistics uncertainty reported from light curve fitting are under-reported, not Gaussian and equation \eqref{eq:pop} is violated. We address this in our analysis in two ways - increasing the reported uncertainty, and allowing for a global shift in observed colour.}

We model the extra dispersion only in colour, and do so by adding extra independent uncertainty on the colour observation. As shown in \citep{Kessler2013}, the extra colour dispersion shows heavy redshift dependence. As the extra dispersion may arise from underlying supernova physics, rather than Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to sufficient extra uncertainty in our model. The $\kappa$ parameters are tightly constrained as they are highly degenerate with the width of the intrinsic colour population $\sigma_c$, and the specific value of this prior was determined from running many iterations of simulation fitting with differing prior strengths on our simple data set with simulated extra dispersion according to the results of \citet{Kessler2013}. The value $0.002$ can be varied by a factor of approximately three in either direction whilst giving stable results, with larger values (more than a factor of three) for the prior resulting in incorrect determination of $\sigma_c$, whilst smaller values resulted in being unable to capture the extra dispersion.  Each survey has its own set of $\kappa$ parameters. As such, our combined covariance on the observation $\hat{\eta}$ is given by $\mathcal{C}_{\rm tot} = \mathcal{C} + {\rm DiagMatrix}\left[0, 0, (\kappa_0 + \kappa_1 z)^2\right]$.

@@

Our second way of addressing the issues with under-reported uncertainty and biased summary statistics are to use fiducial simulations (discussed further in Section \ref{sec:selection}) to characterise the shift in the mean colour measured as a function of redshift. {\green This redshift-dependent array is added to our data-set and is included in our modelling by allowing a controlled subtraction of this shift from our observed data, where the amount of shift subtracted is dictated by parameter $s_c$. Specifically, allow the observed colour to shift by $c_{\rm new} = c_{\rm biased} - s_c \times {\rm bias}(z)$. Unlike $\kappa_0$ and $\kappa_1$, which are redefined for each survey, $s_c$ represents a global shift in supernovae colour across all surveys. We place a uniform prior on $s_c$.} With the extra dispersion introduced from the $\kappa$ parameters addressing the under-reporting of error, shifting the mean colour address the non-Gaussianity introduced by chromatic smearing. We allow the amount of shift to be parametrised to account for different intrinsic scatter models where parameter $s_c$ controls the fraction of bias subtraction -- 0 for no shift, up to 1 for the maximum shift found in simulations. The scatter models, sourced from \citet{Guy2010} and \citet{Chotard2011} are discussed in detail in Section \ref{sec:simdes}. The maximum shift was found to using the intrinsic scatter model from \citet{Chotard2011}. It is important to note that the effects being modelled and characterised by simulations here are features of the supernova population, not cosmological features, and so the choice of fiducial cosmology has negligible impact.

@@

	% Plates
	\plated[thick] {sn} {(obssum)(zobs)(mobs)(pobs)(sum)(z)(m)(mu)} {$\forall\ \rm{SN}$} ;
	\plated[thick] {survey} {(sn)(pop1)} {\\$\forall\ \rm{Survey}$} ;
	\end{tikzpicture}
	\caption{Probabilistic graphical model for our likelihood. Double-lined nodes represent observed variables, diamond nodes represent deterministic variables, and standard nodes represent fit variables. The SN box represents observed variables and latent variables for each individual supernova, whilst the survey box represents survey-specific variables, which in general describe the supernova population for the survey and the systematics associated with it. Variables that appear outside both boxes represent top level model parameters.}
	\label{fig:pgm}
 \end{figure}
